ppseq: An R Package for Sequential Predictive Probability Monitoring

### 2023-04-05 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
fixing bug and deprecated function (#20)
* fixing bug and deprecated function

* Increment version number to 0.2.1.9000

* Update NEWS.md

* updating version

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
abstract: >
  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer. Yet the promise of precision medicine comes with the cost of increased complexity. Recent trials of targeted treatments have included expansion cohorts with sample sizes far exceeding those in traditional early phase trials of chemotherapeutic agents. The enlarged sample sizes raise ethical concerns for patients who enroll in clinical trials, and emphasize the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials of binary endpoints using sequential futility monitoring based on Bayesian predictive probability. Trial designs can be compared using interactive plots and selected based on measures of efficiency or accuracy.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.com/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 60 College Street
    - New Haven, CT 06511
date: "August 20, 2021"
type: package
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article
bibliography: zabor-hobbs-kane.bib
---

@@

Bayesian predictive probability has been proposed as an approach for sequential monitoring in early phase oncology trials [@Dmitrienko2006; @Lee2008; @Hobbs2018; @Saville2014]. However, in order to be useful to investigators designing such trials, software must be made available to calibrate the design for the desired statistical properties. To our knowledge no such software currently exists in the R programming language. This paper introduces the \CRANpkg{ppseq} package for the R software language [@RCT2020], which provides functions to design early phase clinical trials of binary endpoints using sequential predictive probability monitoring for futility. Static plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and interactive plots produced using the \CRANpkg{plotly} package [@Sievert2020] compare design options based on frequentist type I error and power. Moreover, the \CRANpkg{ppseq} package implements two criteria for selecting an ideal predictive probability monitoring design from among the options formed by combinations of posterior and predictive decision thresholds. While the \CRANpkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential futility monitoring in clinical trial design.

@@

```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4.4, out.width = "110%", include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with the {ggplot2} package. The accuracy designs are on the left and the efficiency designs are on the right. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."}
p <- plot(
  two_sample_cal_tbl,
  type1_range = c(0.05, 0.1),
  minimum_power = 0.7,
  plotly = FALSE
  )

@@

```{r echo = FALSE, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of efficiency design options made with the {plotly} package. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond. A similar plot is available for the accuracy design options."}
p <- plot(
  two_sample_cal_tbl,
  type1_range = c(0.05, 0.1),
  minimum_power = 0.7,
  plotly = FALSE
  )
p2_v2 <-
  p[[2]] +
  scale_x_continuous(limits = c(26, 30)) + 
  scale_y_continuous(limits = c(43, 47))

@@

```{r message = FALSE, fig.height = 6, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot. The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
dtp <- plot(two_sample_decision_tbl, plotly = FALSE)

dtp + 
  theme(legend.position="bottom")
```

```{r message = FALSE, fig.height = 6, fig.width = 8, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of decision rules made with plotly. The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
plot(two_sample_decision_tbl)
```

>>>>

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
abstract: >
  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer. Yet the promise of precision medicine comes with the cost of increased complexity. Recent trials of targeted treatments have included expansion cohorts with sample sizes far exceeding those in traditional early phase trials of chemotherapeutic agents. The enlarged sample sizes raise ethical concerns for patients who enroll in clinical trials, and emphasize the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials of binary endpoints using sequential futility monitoring based on Bayesian predictive probability. Trial designs can be compared using interactive plots and selected based on measures of efficiency or accuracy.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.com/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 60 College Street
    - New Haven, CT 06511
date: "August 20, 2021"
type: package
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
bibliography: zabor-hobbs-kane.bib
---

@@

Bayesian predictive probability has been proposed as an approach for sequential monitoring in early phase oncology trials [@Dmitrienko2006; @Lee2008; @Hobbs2018; @Saville2014]. However, in order to be useful to investigators designing such trials, software must be made available to calibrate the design for the desired statistical properties. To our knowledge no such software currently exists in the R programming language. This paper introduces the \CRANpkg{ppseq} package for the R software language, which provides functions to design early phase clinical trials of binary endpoints using sequential predictive probability monitoring for futility. Static plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and interactive plots produced using the \CRANpkg{plotly} package [@Sievert2020] compare design options based on frequentist type I error and power. Moreover, the \CRANpkg{ppseq} package implements two criteria for selecting an ideal predictive probability monitoring design from among the options formed by combinations of posterior and predictive decision thresholds. While the \CRANpkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential futility monitoring in clinical trial design.

@@

```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4.4, out.width = "110%", include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with the {ggplot2} package. The accuracy designs are on the left, with type I error on the x-axis and power on the y-axis, and the efficiency designs are on the right, with the average sample size under the null on the x-axis and the average sample size under the alternative on the y-axis. Each point represents one potential design based on a specific combination of posterior and predictive thresholds. The color represents the Euclidean distance to the top left point. The optimal design is indicated by a diamond. The optimal accuracy design is the one with posterior threshold 0.86 and predictive threshold 0.1. The optimal efficiency design is the one with posterior threshold 0.92 and predictive threshold 0.05."}
p <- plot(
  two_sample_cal_tbl,
  type1_range = c(0.05, 0.1),
  minimum_power = 0.7,
  plotly = FALSE
  )

@@

```{r echo = FALSE, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of efficiency design options made with the {plotly} package. The average sample size under the null is on the x-axis and the average sample size under the alternative is on the y-axis. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond. The optimal efficiency design is the one with posterior threshold 0.92 and predictive threshold 0.05. A similar plot is available for the accuracy design options."}
p <- plot(
  two_sample_cal_tbl,
  type1_range = c(0.05, 0.1),
  minimum_power = 0.7,
  plotly = FALSE
  )
p2_v2 <-
  p[[2]] +
  scale_x_continuous(limits = c(26, 30)) + 
  scale_y_continuous(limits = c(43, 47))

@@

```{r message = FALSE, fig.height = 6, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with {ggplot2}. Each facet is for the total sample size at a pre-specified interim analysis. The number of responses in the control arm is on the x-axis and the number of responses in the experimental arm is on the y-axis. The fill color indicates whether the trial should stop (purple) or proceed (yellow) for a given combination of number of responses in the two arms at each interim analysis. At a given interim analysis, one can look up the number of observed responses to make a decision about whether the trial should stop or proceed enrollment."}
dtp <- plot(two_sample_decision_tbl, plotly = FALSE)

dtp + 
  theme(legend.position="bottom") + 
  scale_fill_viridis_d()
```

```{r message = FALSE, fig.height = 6, fig.width = 8, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of decision rules made with {plotly}. Each facet is for the total sample size at a pre-specified interim analysis. The number of responses in the control arm is on the x-axis and the number of responses in the experimental arm is on the y-axis. The fill color indicates whether the trial should stop (purple) or proceed (yellow) for a given combination of number of responses in the two arms at each interim analysis. At a given interim analysis, one can look up the number of observed responses to make a decision about whether the trial should stop or proceed enrollment."}
ggplotly(
  plot(two_sample_decision_tbl, plotly = FALSE) + scale_fill_viridis_d()
)
```
